Finiteness, Cardinality, and Combinatorics in Homotopy Type Theory

### 2020-01-06 paper.tex
shortened up

  We formalise our work in Cubical Agda
  \cite{vezzosiCubicalAgdaDependently2019}, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
\section{Introduction}
Proofs in a constructive setting are more substantial things than their
classical counterparts.
Rather than just evidence for some fact, they are objects in their own right,
which we can manipulate and extract information from.

Sometimes, the extra detail can reveal interesting subtleties: fissures which
divide homogeneous-seeming definitions into a variety of related concepts.
Finiteness is a prime example of this phenomenon: in classical set theory, a
finite set is a finite set, and there's not much else that we can say.
In constructive type theory, we start the bidding at no fewer than \emph{four}
different predicates representing finiteness
\cite{spiwackConstructivelyFinite2010}, upped to eight in
\cite{firsovVariationsNoetherianness2016a}: all of which differ in significant
and interesting ways.

Other times, the extra baggage is unwanted.
Maybe we want to enforce that a particular piece of information should be
hidden, or treated as irrelevant: ``yes, I have proven that there is a
particular number with this property, but I don't want to reveal \emph{which}
number.''

This is where Homotopy Type Theory \cite{hottbook} can step in.
Through primarily a generalisation of equality, we gain access to a number of
exotic types which can, among other things, allow us to perform this ``hiding''.
In the context of finiteness, the act of hiding is worth studying in its own
right, and further delineates several new finiteness predicates which collapse
even in standard Martin-Löf type theory
\cite{martin-lofIntuitionisticTypeTheory1980}.

The other headline feature of HoTT's generalised equality is univalence: put
simply, it allows us to treat isomorphic types as equal, giving us in the
constructive world access to a technique that is commonplace classically.
It turns out that finite types have a lot to say about isomorphisms, and we will
rely heavily on this aspect of HoTT to work with them.
\subsection{Contributions}
In this work, we will explore finite types in Cubical Type Theory
\cite{cohenCubicalTypeTheory2016}, and expose their relationship to infinite
types, species, and demonstrate their practical uses for proofs in dependently
typed programming languages.
\subsubsection{Strong Finiteness Predicates}
We will first explore the ``strong'' notions of finiteness (i.e. those at least
as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
with a special focus on cardinal finiteness (section~\ref{cardinal}), and
manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
our knowledge.

\input{figures/finite-classification}
Figure~\ref{finite-classification} organises the predicates according to
their ``strength''; i.e. how much information they provide about a conforming
type.
For instance, a proof that some type \(A\) is manifestly Bishop finite (the
strongest of the notions, explored in section~\ref{manifest-bishop}) also tells
us that \(A\) is discrete (has decidable equality), and gives us a linear order
on the type.
A type that is Kuratowski finite (section~\ref{kuratowski}) has no such extra
features: indeed, we will see examples of Kuratowski finite types which are not
even sets, never mind discrete ones.

We will go through each of the predicates, proving how to weaken each (i.e. we
will provide a proof that every cardinally finite type is Kuratowski finite),
and how to strengthen them, given the required property.
In terms of figure~\ref{finite-classification}, this amounts to providing proofs
for each arrow.

We will---through the use of containers
\cite{abbottContainersConstructingStrictly2005}---formally prove the equivalence
these predicates have with the usual function relations i.e. we will show that a
proof of manifest enumerability is precisely equivalent to a surjection from a
finite prefix of the natural numbers.

For each predicate, we will also prove its closure over sums and products in
both dependent and non-dependent forms, if such a closure exists.
This will culminate in our main result for this section: the formal proof that
decidable Kuratowski finite sets form a topos. \todo[noline]{reference for this}
More specifically, we show that cardinal finite sets form a topos, that
decidable Kuratowski finite sets are equivalent in strength to cardinal finite
sets, and carry the proof over.
This result relies on proofs on each of the other finiteness predicates.
\subsubsection{Species}
\todo[inline]{Redo this next paragraph}
In section~\ref{species-and-containers}, we will redefine our finiteness
predicates using \emph{containers}, and prove that the new definitions are
equivalent to the old.
Using this, we will prove much stronger relationships between our finiteness
predicates and relation-based definitions of finiteness.
Finally, We will show how the species-container duality falls out naturally from
these new definitions.

\subsubsection{Infinite Types}
In section~\ref{infinite-cardinalities}, we will extend our study of finite
types to infinite but countable types.
We will see that the finiteness predicates are mirrored with countable
counterparts, and we will prove closure under the Kleene star and plus.

\subsubsection{Practical Uses of Finiteness}
Proofs of finiteness have well-known practical applications in
constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
In section~\ref{practical}, we build a library which exploits these uses in
Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
complex proofs over finite types.
We frame this in terms of the principle of omniscience for finite types.
Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
go further than the usual examples of this kind of proof automation: as well as
proving properties about functions, we can synthesise functions whole-cloth from
their desired properties.
Through the unified interface for finite and countable types, we can
reuse the automation machinery for \emph{partial} proof search over infinite
search spaces.
Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
prove isomorphisms between the curried and uncurried forms of \(n\)-ary
dependent functions.
\section{Split Enumerability} \label{split-enumerable}
We will start with the simplest definition of finiteness: we say a set is
enumerable if there is a list of its elements which contains every element in
the set.
More formally:
\begin{rm-definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
\end{rm-definition}

@@

\begin{rm-definition}[List Membership] \label{list-membership}
  \begin{equation}
    x \in \mathit{xs} = \sum_{(i : \mathbf{Fin}(\text{length}(\mathit{xs})))} , \mathit{xs} \ind i \equiv x
  \end{equation}
  Here, \(\ind\) is a safe indexing function.
  In other words, to say that some item \(x\) is in some list \(\mathit{xs}\),
  we say that there is an index into \(\mathit{xs}\) which selects an element of
  \(\mathit{xs}\) which is equivalent to \(x\).
\end{rm-definition}
\begin{rm-definition}[\mathbf{Fin}]
  \(\mathbf{Fin}\)(n) is the type of natural numbers smaller than \(n\).
  It is defined inductively, as follows:
  \begin{equation}
    \begin{aligned}
      \mathbf{Fin}(0) &= \bot \\
      \mathbf{Fin}(n + 1) &= 1 + \mathbf{Fin}(n)
    \end{aligned}
  \end{equation}
\end{rm-definition}

@@

\subsection{Split Surjections}
Another, equivalent way to define ``finiteness'' is via a (split) surjection
from a finite prefix of the natural numbers.
In this section, we will prove that equivalence, formally.

\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  This proof is the composition of two intermediate equivalences: lemmas
  \ref{container-split-enum-equiv} and \ref{container-split-enum-surj}.
  This first of these proves that our definition of finiteness is equivalent to
  another predicate which uses lists defined as a \emph{container}
  (definition~\ref{container-def}).
  The second of these shows that the container-based finiteness predicate is
  equivalent to a split surjection from some \(\mathbf{Fin}\).
\end{proof}

Our interest in the container-based finiteness predicate comes from the fact
that it is very straightforward to prove it equivalent to a split surjection
from some \(\mathbf{Fin}\).
\begin{rm-lemma} \label{container-split-enum-surj}
  Container-defined split enumerability is equivalent to a split surjection from
  some \(\mathbf{Fin}\).
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-lemma}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
\end{proof}
\subsection{Decidable Equality}
\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
\begin{rm-lemma}
  Any split enumerable type is a set.
\end{rm-lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma}
  \(\bot\) is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support list for \(\bot\) is the empty list.
  The cover proof is present via the principle of explosion.
\end{proof}
\begin{rm-lemma}
  \(\top\) is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support list for \(\top\) is the singleton list containing \(\text{tt}\).
  The cover proof is a constant function which returns an index into the first
  item in the list.
\end{proof}
Both of those cases were relatively simple.
Next, we will look into how to combine proofs of split enumerability.
\begin{rm-theorem}
  Split-enumerability is closed under \(\sum\).
\end{rm-theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \prod_{(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

@@

  The cover proof follows in a simple, if tedious, way.
\end{proof}
\begin{rm-lemma}
  Split-enumerability is closed under Cartesian product (non-dependent product).
\end{rm-lemma} 
\begin{proof}
  Since non-dependent products are simply a special case of the \(\sum\) type,
  we can reuse the proof of closure over \(\sum\) here.
\end{proof}
\begin{rm-lemma}
  Split-enumerability is closed under disjoint union (non-dependent sum).
\end{rm-lemma}
\begin{proof}
  Again, this can be derived from closure of the \(\sum\) type.
  Disjoint union between two types \(A\) and \(B\) can be represented by the
  following type:
  \begin{equation}
    A + B = \sum_{(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
  \end{equation}
  Then, since all of \(\mathbf{Bool}\), \(A\), and \(B\) are split enumerable,
  the type \(A + B\) is split enumerable.
\end{proof}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
\begin{rm-definition}[Manifest Bishop Finiteness]
  \begin{equation}
    \mathcal{B}(A) = \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \inunique xs
  \end{equation}
\end{rm-definition}
The only difference between this predicate and split enumerability is the list
membership term: we use \(\inunique\) here, where \(x \inunique \mathit{xs}\) is
to be read as ``\(x\) occurs exactly once in \(\mathit{xs}\)''.
It is defined in terms of our previous list membership term
(definition~\ref{list-membership}).
\begin{rm-definition}[Unique List Membership]
  We say an item \(x\) is ``uniquely in'' some list \(\mathit{xs}\) if its
  membership in that list is a \emph{contraction}; i.e. its membership proof
  exists, and all such proofs are equal.
  \begin{equation}
    x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{rm-definition}

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsection{Equivalence}
While manifest Bishop finiteness may seem stricter than split enumerability, we
will see soon that it is actually equivalent in strength.
Why study it separately to split enumerability at all, then?
What purpose does it serve on our road to Kuratowski finiteness?
Those questions are answered by the following theorem:
\begin{rm-lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{rm-lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  In other words, we can define a container-based version of bishop finiteness
  which is a reassociation of a contractible map, and then show that the
  container-based predicate is equivalent to the list-based.
\end{proof}
\subsection{Relationship to Split Enumerability}
We now show that manifest Bishop finiteness has equal strength to split
enumerability.
\begin{rm-lemma}
  Any manifest Bishop finite set is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support set carries over simply, and the cover proof can be taken from the
  first component of the cover proof from the proof of manifest Bishop
  finiteness.
  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{bishop-to-split-enum}
\end{proof}
\begin{rm-theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{rm-theorem}
\begin{proof}
  Let \(E\) be a proof of split enumerability for some set \(A\).
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function \(\text{uniques}\) which filters
  out duplicates from lists of \(A\)s.
  \begin{equation}
    \text{uniques} : \textbf{List}(A) \rightarrow \textbf{List}(A)
  \end{equation}
  This gives us our support list.

@@

\begin{rm-theorem}
  Any cardinal finite type with a (decidable) total order is manifestly Bishop
  finite.
\end{rm-theorem}
\begin{proof}
  This proof is quite involved, and will rely on several subsequent lemmas, so
  we will give only its outline here.
  \begin{itemize}
    \item First, we will convert to manifest enumerability: knowing that the
      underlying type is discrete (theorem~\ref{cardinal-finite-discrete}) we can
      go from manifest enumerability to split enumerability
      (lemma~\ref{manifest-enum-to-split}), and subsequently to manifest Bishop
      finiteness (lemma~\ref{split-enum-to-manifest-bishop}).
    \item To convert to manifest enumerability, we need to provide a support
      list: this cannot simply be the support list hidden under the truncation,
      since that would violate the hiding promised by the truncation.
      Instead, we sort the list (using insertion sort).
      We must, therefore, prove that insertion sort is invariant under all
      support lists in cardinal finiteness proofs.
    \item We show that all support lists in cardinal finiteness proofs are
      permutations of each other.
    \item And then we show that insertion sort is invariant under permutations.
  \end{itemize} 
\end{proof}
Now we will build up the toolkit we need to perform the above steps.
First, permutations.
\begin{rm-definition}[List Permutations]
  We say that two lists are permutations of each other if there is an
  isomorphism between membership proofs.
  \begin{equation}
    \mathit{xs} \leftrightsquigarrow \mathit{ys} = \prod_{(x : A)} , x \in \mathit{xs} \iff x \in \mathit{ys}
  \end{equation}
\end{rm-definition}
We also prove some of the identities you might expect with regards to
permutations, all in the appendix.
In particular, we prove that they form an equivalence relation.

>>>>

  We formalise our work in Cubical Agda
  \cite{vezzosiCubicalAgdaDependently2019}, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
% \section{Introduction}
% Proofs in a constructive setting are more substantial things than their
% classical counterparts.
% Rather than just evidence for some fact, they are objects in their own right,
% which we can manipulate and extract information from.

% Sometimes, the extra detail can reveal interesting subtleties: fissures which
% divide homogeneous-seeming definitions into a variety of related concepts.
% Finiteness is a prime example of this phenomenon: in classical set theory, a
% finite set is a finite set, and there's not much else that we can say.
% In constructive type theory, we start the bidding at no fewer than \emph{four}
% different predicates representing finiteness
% \cite{spiwackConstructivelyFinite2010}, upped to eight in
% \cite{firsovVariationsNoetherianness2016a}: all of which differ in significant
% and interesting ways.

% Other times, the extra baggage is unwanted.
% Maybe we want to enforce that a particular piece of information should be
% hidden, or treated as irrelevant: ``yes, I have proven that there is a
% particular number with this property, but I don't want to reveal \emph{which}
% number.''

% This is where Homotopy Type Theory \cite{hottbook} can step in.
% Through primarily a generalisation of equality, we gain access to a number of
% exotic types which can, among other things, allow us to perform this ``hiding''.
% In the context of finiteness, the act of hiding is worth studying in its own
% right, and further delineates several new finiteness predicates which collapse
% even in standard Martin-Löf type theory
% \cite{martin-lofIntuitionisticTypeTheory1980}.

% The other headline feature of HoTT's generalised equality is univalence: put
% simply, it allows us to treat isomorphic types as equal, giving us in the
% constructive world access to a technique that is commonplace classically.
% It turns out that finite types have a lot to say about isomorphisms, and we will
% rely heavily on this aspect of HoTT to work with them.
% \subsection{Contributions}
% In this work, we will explore finite types in Cubical Type Theory
% \cite{cohenCubicalTypeTheory2016}, and expose their relationship to infinite
% types, species, and demonstrate their practical uses for proofs in dependently
% typed programming languages.
% \subsubsection{Strong Finiteness Predicates}
% We will first explore the ``strong'' notions of finiteness (i.e. those at least
% as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
% with a special focus on cardinal finiteness (section~\ref{cardinal}), and
% manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
% our knowledge.

% \input{figures/finite-classification}
% Figure~\ref{finite-classification} organises the predicates according to
% their ``strength''; i.e. how much information they provide about a conforming
% type.
% For instance, a proof that some type \(A\) is manifestly Bishop finite (the
% strongest of the notions, explored in section~\ref{manifest-bishop}) also tells
% us that \(A\) is discrete (has decidable equality), and gives us a linear order
% on the type.
% A type that is Kuratowski finite (section~\ref{kuratowski}) has no such extra
% features: indeed, we will see examples of Kuratowski finite types which are not
% even sets, never mind discrete ones.

% We will go through each of the predicates, proving how to weaken each (i.e. we
% will provide a proof that every cardinally finite type is Kuratowski finite),
% and how to strengthen them, given the required property.
% In terms of figure~\ref{finite-classification}, this amounts to providing proofs
% for each arrow.

% We will---through the use of containers
% \cite{abbottContainersConstructingStrictly2005}---formally prove the equivalence
% these predicates have with the usual function relations i.e. we will show that a
% proof of manifest enumerability is precisely equivalent to a surjection from a
% finite prefix of the natural numbers.

% For each predicate, we will also prove its closure over sums and products in
% both dependent and non-dependent forms, if such a closure exists.
% This will culminate in our main result for this section: the formal proof that
% decidable Kuratowski finite sets form a topos. \todo[noline]{reference for this}
% More specifically, we show that cardinal finite sets form a topos, that
% decidable Kuratowski finite sets are equivalent in strength to cardinal finite
% sets, and carry the proof over.
% This result relies on proofs on each of the other finiteness predicates.
% \subsubsection{Species}
% \todo[inline]{Redo this next paragraph}
% In section~\ref{species-and-containers}, we will redefine our finiteness
% predicates using \emph{containers}, and prove that the new definitions are
% equivalent to the old.
% Using this, we will prove much stronger relationships between our finiteness
% predicates and relation-based definitions of finiteness.
% Finally, We will show how the species-container duality falls out naturally from
% these new definitions.

% \subsubsection{Infinite Types}
% In section~\ref{infinite-cardinalities}, we will extend our study of finite
% types to infinite but countable types.
% We will see that the finiteness predicates are mirrored with countable
% counterparts, and we will prove closure under the Kleene star and plus.

% \subsubsection{Practical Uses of Finiteness}
% Proofs of finiteness have well-known practical applications in
% constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
% In section~\ref{practical}, we build a library which exploits these uses in
% Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
% complex proofs over finite types.
% We frame this in terms of the principle of omniscience for finite types.
% Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
% go further than the usual examples of this kind of proof automation: as well as
% proving properties about functions, we can synthesise functions whole-cloth from
% their desired properties.
% Through the unified interface for finite and countable types, we can
% reuse the automation machinery for \emph{partial} proof search over infinite
% search spaces.
% Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
% prove isomorphisms between the curried and uncurried forms of \(n\)-ary
% dependent functions.
\section{Split Enumerability} \label{split-enumerable}
We will start with the simplest definition of finiteness: we say a set is
enumerable if there is a list of its elements which contains every element in
the set.
More formally:
\begin{rm-definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
\end{rm-definition}

@@

\begin{rm-definition}[List Membership] \label{list-membership}
  \begin{equation}
    x \in \mathit{xs} = \sum_{(i : \mathbf{Fin}(\text{length}(\mathit{xs})))} , \mathit{xs} \ind i \equiv x
  \end{equation}
  Here, \(\ind\) is a safe indexing function.
  In other words, to say that some item \(x\) is in some list \(\mathit{xs}\),
  we say that there is an index into \(\mathit{xs}\) which selects an element of
  \(\mathit{xs}\) which is equivalent to \(x\).
\end{rm-definition}
\begin{rm-definition}[\mathbf{Fin}]
  \(\mathbf{Fin}\)(n) is the type of natural numbers smaller than \(n\).
  It is defined inductively, as follows:
  \begin{equation}
    \begin{aligned}
      \mathbf{Fin}(0) &= \bot \\
      \mathbf{Fin}(n + 1) &= \top + \mathbf{Fin}(n)
    \end{aligned}
  \end{equation}
\end{rm-definition}

@@

For each finiteness predicate, we have proven it equivalent to its
container-based version.
These proofs are included in the Agda formalisation.
\subsection{Split Surjections}
Another, equivalent way to define ``finiteness'' is via a (split) surjection
from a finite prefix of the natural numbers.
In this section, we will prove that equivalence, formally.

\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  For this proof, we will work with the container-based finiteness predicate.
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
\end{proof}
\subsection{Decidable Equality}
\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
\begin{rm-lemma}
  Any split enumerable type is a set.
\end{rm-lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma}
  \(\bot\) is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support list for \(\bot\) is the empty list.
  The cover proof is present via the principle of explosion.
\end{proof}
\begin{rm-lemma}
  \(\top\) is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support list for \(\top\) is the singleton list containing \(\text{tt}\).
  The cover proof is a constant function which returns an index into the first
  item in the list.
\end{proof}
Both of those cases were relatively simple.
Next, we will look into how to combine proofs of split enumerability.
\begin{rm-theorem}
  Split-enumerability is closed under \(\sum\).
\end{rm-theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \prod_{(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

@@

  The cover proof follows in a simple, if tedious, way.
\end{proof}
\begin{rm-lemma}
  Split-enumerability is closed under Cartesian product (non-dependent product).
\end{rm-lemma} 
\begin{proof}
\end{proof}
\begin{rm-lemma}
  Split-enumerability is closed under disjoint union (non-dependent sum) and
  Cartesian product (non-dependent product).
\end{rm-lemma}
\begin{proof}
  Both of these closures can be derived from closure of the \(\sum\) type.
  The non-dependent product is equivalent to a special case of \(\sum\):
  \begin{equation}
    A \times B \simeq \sum_{(x : A)} , B
  \end{equation}
  And disjoint union between two types \(A\) and \(B\) can be represented by the
  following type:
  \begin{equation}
    A + B = \sum_{(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
  \end{equation}
  Then, since all of \(\mathbf{Bool}\), \(A\), and \(B\) are split enumerable,
  the type \(A + B\) is split enumerable.
\end{proof}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
\begin{rm-definition}[Manifest Bishop Finiteness]
  \begin{equation}
    \mathcal{B}(A) = \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \inunique xs
  \end{equation}
\end{rm-definition}
The only difference between this predicate and split enumerability is the list
membership term: we use \(\inunique\) here, where \(x \inunique \mathit{xs}\) is
to be read as ``\(x\) occurs exactly once in \(\mathit{xs}\)''.
It is defined in terms of our previous list membership term
(definition~\ref{list-membership}).
\begin{rm-definition}[Unique List Membership]
  We say an item \(x\) is ``uniquely in'' some list \(\mathit{xs}\) if its
  membership in that list is a \emph{contraction}; i.e. its membership proof
  exists, and all such proofs are equal.
  \begin{equation}
    x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{rm-definition}

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsection{Equivalence}
Where split enumerability was the enumeration form of a surjection from
\(\mathbf{Fin}\), we see here that manifest Bishop finiteness is the enumeration
form of an \emph{equivalence} with \(\mathbf{Fin}\).
\begin{rm-lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{rm-lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  In other words, we can define a container-based version of bishop finiteness
  which is a reassociation of a contractible map, and then show that the
  container-based predicate is equivalent to the list-based.
\end{proof}
\subsection{Relationship to Split Enumerability}
We now show that manifest Bishop finiteness has equal strength to split
enumerability.
\begin{rm-lemma}
  Any manifest Bishop finite set is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support set carries over simply, and the cover proof can be taken from the
  first component of the cover proof from the proof of manifest Bishop
  finiteness.
\end{proof}
\begin{rm-theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{rm-theorem}
\begin{proof}
  Let \(E\) be a proof of split enumerability for some set \(A\).
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function \(\text{uniques}\) which filters
  out duplicates from lists of \(A\)s.
  \begin{equation}
    \text{uniques} : \textbf{List}(A) \rightarrow \textbf{List}(A)
  \end{equation}
  This gives us our support list.

@@

\begin{rm-theorem}
  Any cardinal finite type with a (decidable) total order is manifestly Bishop
  finite.
  This proof is quite involved, and will rely on several subsequent lemmas, so
  we will give only its outline here.
  \begin{itemize}
    \item First, we will convert to manifest enumerability: knowing that the
      underlying type is discrete (theorem~\ref{cardinal-finite-discrete}) we can
      go from manifest enumerability to split enumerability
      (lemma~\ref{manifest-enum-to-split}), and subsequently to manifest Bishop
      finiteness (lemma~\ref{split-enum-to-manifest-bishop}).
    \item To convert to manifest enumerability, we need to provide a support
      list: this cannot simply be the support list hidden under the truncation,
      since that would violate the hiding promised by the truncation.
      Instead, we sort the list (using insertion sort).
      We must, therefore, prove that insertion sort is invariant under all
      support lists in cardinal finiteness proofs.
    \item We show that all support lists in cardinal finiteness proofs are
      permutations of each other.
    \item And then we show that insertion sort is invariant under permutations.
  \end{itemize} 
\end{proof}
Now we will build up the toolkit we need to perform the above steps.
First, permutations.
\begin{rm-definition}[List Permutations]
  We say that two lists are permutations of each other if there is an
  isomorphism between membership proofs
  \cite{danielssonBagEquivalenceProofRelevant2012}.
    \mathit{xs} \leftrightsquigarrow \mathit{ys} = \prod_{(x : A)} , x \in \mathit{xs} \iff x \in \mathit{ys}
We also prove some of the identities you might expect with regards to
permutations, all in the appendix.
In particular, we prove that they form an equivalence relation.
\begin{rm-lemma}
  Insertion sort is invariant under permutations.
    \mathit{xs} \leftrightsquigarrow \mathit{ys} \implies \text{sort}(\mathit{xs}) \equiv \text{sort}(\mathit{ys})
\end{rm-lemma}
  We first prove that insertion sort does indeed sort its input: i.e. it returns
  a sorted list that is a permutation of its input.
  Then we show that any two sorted lists which are permutations of each other
  are equal.
\end{proof}
\todo[inline]{Agda code for proofs}
\section{Kuratowski Finiteness} \label{kuratowski}
Finally we arrive at Kuratowski finiteness
\cite{kuratowskiNotionEnsembleFini1920}.
\subsection{The Kuratowski Set as a Higher Inductive Type}
We start with the Kuratowski finite sets, which are free join semilattices (or,
equivalently, free commutative idempotent monoids).
HITs are required to define this type
\cite{altenkirchDefinableQuotientsType2011}; however we have two possible
candidates for the definition.
The first is a direct translation of ``free commutative idempotent monoid'' into
a type:
\begin{aligned}
  \mathcal{K}(A) &=  \\
    &\left\vert \; \text{singleton} : A \rightarrow \mathcal{K}(A) \right; \\
    &\left\vert \; \cdot \cup \cdot : \mathcal{K}(A) \times \mathcal{K}(A) \rightarrow \mathcal{K}(A) \right; \\
    &\left\vert \; \emptyset : \mathcal{K}(A) \right; \\
    &\left\vert \; \cup \text{-assoc} : \Pi \left( \mathit{xs}, \mathit{ys}, \mathit{zs}:\mathcal{K}(A)\right) , (\mathit{xs} \cup \mathit{ys}) \cup \mathit{zs} \equiv \mathit{xs} \cup (\mathit{ys} \cup \mathit{zs}) \right; \\
    &\left\vert \; \cup \text{-commutative} : \Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \mathit{xs} \cup \mathit{ys} \equiv \mathit{ys} \cup \mathit{xs} \right; \\
    &\left\vert \; \cup \text{-idempotent} : \Pi (\mathit{xs}: \mathcal{K}(A)) , \mathit{xs} \cup \mathit{xs} \equiv \mathit{xs} \right; \\
    &\left\vert \; \cup \text{-identity} : \Pi (\mathit{xs}: \mathcal{K}(A)) , \emptyset \cup \mathit{xs} \equiv \mathit{xs} \right; \\
    &\left\vert \; \text{trunc} : \Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \Pi (p, q : \mathit{xs} \equiv \mathit{ys}) , p \equiv q \right;
\end{aligned}
\todo[inline]{Have we already described the trunc constructor?}
